---
title: "Cotton fields soil microbiome Project"
author: "Matthew Gillis and Pedro Rodrigues"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    fig_caption: yes

fontsize: 14pt
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

```{r init, warning=FALSE, message=FALSE}
setwd("/Users/pedro/Documents/GitHub/Cotton-Soil-Micro/")
set.seed("12345678")
library(dada2);packageVersion("dada2") # 1.30.0
library(phyloseq);packageVersion("phyloseq") # 1.46.0
library(Biostrings); packageVersion("Biostrings") # 2.70.3
library(rmarkdown); packageVersion("rmarkdown") # 2.26
library(tidyverse); packageVersion("tidyverse") # 2.0.0
```

```{r more preparation, warning=FALSE, message=FALSE}
path.out <- "Results/Figures/"
path.rds <- "Input/RDS/"
theme_set(theme_bw())
```

Load Data (quality-trimmed and denoised using DADA2 and the nf-core/ampliseq pipeline)

```{r phyloseq-load-improved-data, eval=FALSE}
p_data.chimera_free <- readRDS("../Input/RDS/DADA2_table.rds")
```

```{r assign taxonomy, eval=FALSE}
# Assign taxonomy
tax <- assignTaxonomy(p_data.chimera_free, "../Input/rdp_19_toGenus_trainset.fa.gz", minBoot = 80, tryRC = TRUE, multithread=TRUE) # Slowest part
head(unname(tax))
tax <- addSpecies(tax, "../Input/rdp_species_assignment_18.fa.gz")
```

```{r save-progress-taxonomy, eval=FALSE}
# Write to disk
saveRDS(tax, "../Input/RDS/tax_soilmicrobes.rds")
```

Make Phyloseq object

```{r phyloseq-load-data, eval=FALSE}
p_data.chimera_free <- readRDS("../Input/RDS/DADA2_table.rds")
tax <- readRDS("../Input/RDS/tax_soilmicrobes.rds")
metadata <- read.table("../Input/metadata_matthew.txt", sep = "\t", header = TRUE)
samples.out <- data.frame(old_ID = rownames(p_data.chimera_free))
samples_out <- samples.out %>% separate_wider_delim(old_ID,delim ="_1.f", names = c("ID", "suffix"), cols_remove = FALSE)
metadata_merge <- left_join(metadata, samples_out, by = "ID")
row.names(metadata_merge) <- metadata_merge[,6]
```

```{r backup-data, eval=FALSE}
write_tsv(metadata_merge, "../Input/metadataPhyloseqFormatted.tsv")
```


```{r phyloseq-make-object-1, eval=FALSE}
p_data <- phyloseq(otu_table(p_data.chimera_free, taxa_are_rows = FALSE),
                   sample_data(metadata_merge),
                   tax_table(tax))
```

Store DNA sequences in a different slot and simplify ASV names

```{r phyloseq-format-data, eval=FALSE}
dna <- Biostrings::DNAStringSet(taxa_names(p_data))
names(dna) <- taxa_names(p_data)
p_data <- merge_phyloseq(p_data, dna)
taxa_names(p_data) <- paste0("ASV", seq(ntaxa(p_data)))
p_data
```

Investigate taxa in the dataset - how many samples are not bacteria?

```{r summarize-no-bacteria, eval=FALSE}
p_data
p_data %>%
  subset_taxa(Kingdom!="Bacteria") %>%
  tax_table() %>% as.data.frame() %>% group_by(Class) %>% tally()
```

One Plant taxa, and thirty-two Archaea.

How many identified and unidentified ASV at the Phylum and Class levels?

```{r summarize-no-bacteria-phylum-class, eval=FALSE}
p_data
p_data %>%
  tax_table() %>% as.data.frame() %>% group_by(Phylum, Class) %>% tally() %>% paged_table()
```

A few NA. Let's exclude obvious contaminants before saving sequences as a fasta file and using Blast to (hopefully) get higher definition for taxa currently classified as "NA".

First, I will make data easier to work with, by transforming the phyloseq object into a summary table.

```{r relative-abundance-dada2-processed-data-and-taxonomy, eval=FALSE}
pdata_df <- p_data %>% psmelt() %>% arrange(OTU) %>% dplyr::rename(ASV = OTU) %>% spread(Sample, Abundance)
```

Let's remove chloroplasts, plant DNA and other taxa that may not be relevant here.

```{r remove-chloroplast--plants-archaea, eval=FALSE}
p_data_taxlean <- p_data %>% subset_taxa(Kingdom != "Archaea") %>% subset_taxa((Order != "Chloroplast") | is.na(Order)) %>% subset_taxa((Kingdom != "Eukaryota") | is.na(Kingdom)) %>% subset_taxa((Class != "Cyanobacteria") | is.na(Order)) %>% subset_samples(sample_sums(p_data) > 0)
```

Save taxonomy table and ASV fasta and blast unknown sequences to increase accuracy of classification

```{r save-taxonomy-table, eval=FALSE}
p_data_taxlean %>%
  tax_table() %>%
  write.csv("../Results/MG_asv_04_2025.csv")
```

```{r save-sequences-fasta, eval=FALSE}
p_data_taxlean %>%
      refseq() %>%
      Biostrings::writeXStringSet("../Results/MG_asv_04_2025.fasta", append=FALSE,
                                  compress=FALSE, compression_level=NA, format="fasta")
```

Blast was ran using the script "blast.sh", and taxonomical ranks were added to the output file ("28APR2025_nt_blast_MG.tsv") using the script "Get_lineage_blast.sh". The resulting spreadsheet was saved as "soil_16S_withtaxranks_ASV_nt_blast.csv". Next, I will identify "NA" sequences that were resolved by blast, and manually inspect their identification and eliminate any spurious results (e.g. annotations containing mistakes, non-taxonomical attributions, etc).

```{r load taxonomical data}
old_tax_table <- read.csv("../Results/MG_asv_04_2025.csv")
blast_tax_table <- read.csv("../Results/blast_results/soil_16S_withtaxranks_ASV_nt_blast.csv")
NA_old_tax_table <- old_tax_table %>% filter(is.na(Genus)) %>% 
  rename("qseqid" = "X") #select only ASV with NA at the Genus level; rename ASV id column to match the corresponding blast column name
NoNA_blast_tax_table <- blast_tax_table %>% filter(!is.na(genus)) # select blast results that resolved identification at the Genus level
resolved_by_blast <- left_join(NA_old_tax_table, NoNA_blast_tax_table, by = "qseqid") %>% filter(!is.na(genus)) # combine both subset data sets, using the old tax table as the reference for rows to include, and qseqid as the column in common to allow the merge of these tables; next, exclude "genus" in blast results that were not resolved
```

Blast helped resolve the identification of 214 ASVs. Next, I will transfer all taxonomical levels for resolved genera from blast results to the "old_tax_table".

```{r update taxonomy}
# Prepare resolved_by_blast to be added to the final taxonomy file
# Select only relevant columns
resolved_by_blast_slim <- resolved_by_blast %>% select(c("qseqid", "kingdom", "phylum", "class", "order", "family", "genus", "species"))
# Rename columns to match phyloseq tax table
colnames(resolved_by_blast_slim) <- c("X", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
# Rename values in "Species" to eliminate characters that Phyloseq might not like
resolved_by_blast_slim <- resolved_by_blast_slim %>% mutate(Species = paste ("sp"))
# Use anti-join to eliminate from old_tax_table entries that overlap with the new results
old_tax_table_slice <- anti_join(old_tax_table, resolved_by_blast_slim, by = "X")
# Now join rows from both datasets into the final dataset
updated_tax <- bind_rows(resolved_by_blast_slim, old_tax_table_slice)
```

For now this should be sufficient. Downstream, I will further improve taxonomy by replacing NA at the genus level with the next best taxonomical id followed by "sp" (e.g. "Pseudomonadaceae sp.", or "Bacteria sp.", etc).

Add new taxonomy back to the phyloseq object

```{r add-updated-taxonomy}
#rename rows with ASV identifier
row.names(updated_tax) <- updated_tax$X
# remove first column
updated_tax <- updated_tax[-1]
new_tax <- tax_table(as.matrix(updated_tax))
tax_table(p_data_taxlean) <- new_tax
```